---
title: "Seeking Help in R"
teaching: 10
exercises: 10
questions:
- "How can I get help in R?"
objectives:
- "To be able read R help files for functions and special operators."
- "To be able to use CRAN task views to identify packages to solve a problem."
keypoints:
- "Use `help()` to get online help in R."
source: Rmd
---

```{r, include=FALSE}
source("bin/chunk-options.R")
knitr_fig_path("02-")
```

# Seeking Help in R

## Reading Help files

R, and every package, provide help files for functions. The general syntax to search for help on any
function, "function_name", from a specific function that is in a package loaded into your
namespace (your interactive R session):

```{r, eval=FALSE}
?function_name
help(function_name)
```

This will load up a help page in RStudio (or as plain text in R by itself).

Each help page is broken down into sections:

 - Description: An extended description of what the function does.
 - Usage: The arguments of the function and their default values.
 - Arguments: An explanation of the data each argument is expecting.
 - Details: Any important details to be aware of.
 - Value: The data the function returns.
 - See Also: Any related functions you might find useful.
 - Examples: Some examples for how to use the function.

Different functions might have different sections, but these are the main ones you should be aware of.

> ## Tip: Running Examples
>
> From within the function help page, you can highlight code in the 
> Examples and hit <kbd>Ctrl</kbd>+<kbd>Return</kbd> to run it in 
> RStudio console. This is gives you a quick way to get a feel for
> how a function works.

> ## Tip: Reading help files
>
> One of the most daunting aspects of R is the large number of functions
> available. It would be prohibitive, if not impossible to remember the
> correct usage for every function you use. Luckily, the help files
> mean you don't have to!


## Special Operators

To seek help on special operators, use quotes:

```{r, eval=FALSE}
?"<-"
```

## Getting help on packages

Many packages come with "vignettes": tutorials and extended example documentation.
Without any arguments, `vignette()` will list all vignettes for all installed packages;
`vignette(package="package-name")` will list all available vignettes for
`package-name`, and `vignette("vignette-name")` will open the specified vignette.

If a package doesn't have any vignettes, you can usually find help by typing
`help("package-name")`.

Many packages also have a web presence. Google is your friend here. For example here is the package website for the ggplot2 package: https://ggplot2.tidyverse.org/

## When you kind of remember the function

If you're not sure what package a function is in, or how it's specifically spelled you can do a fuzzy search:

```{r, eval=FALSE}
??function_name
```

## When you have no idea where to begin

If you don't know what function or package you need to use
[CRAN Task Views](http://cran.at.r-project.org/web/views)
is a specially maintained list of packages grouped into
fields. This can be a good starting point.

### Challenge 1

> Look at the help for the `sum` function. What are two ways you can pass numbers 
> into the function so the are added together?
> 
> <details>
> 
> <summary>
> Solution to challenge 1
> </summary>
> 
> <br />
> 
> ```{r, eval=FALSE}
> ?sum
> sum(1, 2, 3, 4, 5)
> x <- c(1, 2, 3)
> sum(x)
> ```
> 
> </details>




### Challenge 2

> Look at the help for the `paste` function. You'll need to use this later.
> What is the difference between the `sep` and `collapse` arguments?
> 
> 
> <details>
> 
> <summary>
> Solution to challenge 2
> </summary>
> 
> <br />
> 
> To look at the help for the `paste()` function, use:
> ```{r, eval=FALSE}
> help("paste")
> ?paste
> ```
> The difference between `sep` and `collapse` is a little
> tricky. The `paste` function accepts any number of arguments, each of which
> can be a vector of any length. The `sep` argument specifies the string
> used between concatenated terms — by default, a space. The result is a
> vector as long as the longest argument supplied to `paste`. In contrast,
> `collapse` specifies that after concatenation the elements are *collapsed*
> together using the given separator, the result being a single string.
> e.g. 
> ```{r} 
> paste(c("a","b"), "c")
> paste(c("a","b"), "c", sep = ",")
> paste(c("a","b"), "c", collapse = "|")
> paste(c("a","b"), "c", sep = ",", collapse = "|")
> ```
> (For more information,
> scroll to the bottom of the `?paste` help page and look at the
> examples, or try `example('paste')`.)
> 
> </details>




### Challenge 3
> Use help to find a function (and its associated parameters) that you could
> use to load data from a csv file in which columns are delimited with "\t"
> (tab) and the decimal point is a "." (period). This check for decimal
> separator is important, especially if you are working with international
> colleagues, because different countries have different conventions for the
> decimal point (i.e. comma vs period).
> hint: use `??csv` to lookup csv related functions.
> 
> 
> <details>
> 
> <summary>
> Solution to challenge 3
> </summary>
> 
> <br />
> 
> The standard R function for reading tab-delimited files with a period
> decimal separator is read.delim(). You can also do this with
> `read.table(file, sep="\t")` (the period is the *default* decimal
> separator for `read.table()`, although you may have to change
> the `comment.char` argument as well if your data file contains
> hash (#) characters
> 
> </details>



