---
title: "Exploring Data Frames"
teaching: 20
exercises: 10
questions: "How can I manipulate a data frame?"
objectives:
- "Display basic properties of data frames including size and class of the columns, names, and first few rows."
keypoints:
- "Use `str()`, `summary()`, `nrow()`, `ncol()`, `dim()`, `colnames()`, `rownames()`, `head()`, and `typeof()` to understand the structure of a data frame."
- "Read in a csv file using `readr::read_csv()`."
source: Rmd
---

# Exploring Data Frames

```{r, include=FALSE}
source("bin/chunk-options.R")
knitr_fig_path("05-")
```

At this point, you've seen it all: in the last lesson, we toured all the basic
data types and data structures in R. Everything you do will be a manipulation of
those tools. But most of the time, the star of the show is the data frame—the table that we created by loading information from a csv file. In this lesson, we'll learn a few more things
about working with data frames.

```{r}
library(readr)
cats <- read_csv("data/feline-data.csv")
```

We can find the size and various other attributes of our data with a few handy functions:

```{r}
nrow(cats)
ncol(cats)
dim(cats)
```

```{r}
names(cats)
str(cats)
```

A way to get a nice overall view of a data frame is to use the `summary` function. This function can be used on various objects in R. For data frames, `summary` yields a numeric, tabular, or descriptive summary of each column. Factor columns are summarized by the number of items in each level, numeric or integer columns by the descriptive statistics (quartiles and mean), and character columns by its length, class, and mode.  

```{r}
summary(cats)
```

A really nice feature of RStudio is that it allows us to view our data in a 
very readable and familiar way. Go ahead and click on the name of the `cats`
data frame in the environment pane.

Note that this is a read-only view, it's not editable (which is a good thing!)

## Introducing the Gapminder data

So far, you have seen the basics of manipulating data frames with our cat data;
now let's use those skills to digest a more realistic dataset. Let's read in the
`gapminder` dataset that we downloaded previously and assign it to a data frame called `gapminder`:

 ```{r}
 gapminder <- read_csv("data/gapminder_data.csv")
 ```

### Challenge 1 (5 minutes)
>
> How many rows and columns does `gapminder` have?
>
> What kind of data is in the `pop` and `continent` columns?
>
> <details>
> 
> <summary>
> Solution to challenge 1
> </summary>
> 
> <br />
> 
> ```{r}
> dim(gapminder)
> nrow(gapminder)
> ncol(gapminder)
> str(gapminder)
> summary(gapminder)
> ```
> </details>

The first thing we should always do is check out what the data looks like with
`str()` or `summary()`:

```{r}
str(gapminder)
```


```{r}
summary(gapminder$country)
```

::: {.rmdtip}

**Miscellaneous Tips**

* Files can also be downloaded directly from the Internet into a local
folder of your choice onto your computer using the `download.file` function.
The `read_csv` function can then be executed to read the downloaded file from the download location, for example,
```{r eval=FALSE, echo=TRUE}
download.file("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv", destfile = "data/gapminder_data.csv")
gapminder <- read_csv("data/gapminder_data.csv")
```

* Alternatively, you can also read in files directly into R from the Internet by replacing the file paths with a web address in `read_csv`. One should note that in doing this no local copy of the csv file is first saved onto your computer. For example,
```{r eval=FALSE, echo=TRUE}
gapminder <- read_csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv")
```

* You can read directly from excel spreadsheets without
converting them to plain text first by using the [readxl](https://cran.r-project.org/package=readxl) package.

:::


We'll also likely want to know what the titles of all the columns are, so we can
ask for them:

```{r}
names(gapminder)
```

At this stage, it's important to ask ourselves if the structure R is reporting
matches our intuition or expectations; do the basic data types reported for each
column make sense? If not, we need to sort any problems out now before they turn
into bad surprises down the road, using what we've learned about how R
interprets data, and the importance of *strict consistency* in how we record our
data.

Once we're happy that the data types and structures seem reasonable, it's time
to start digging into our data proper. Check out the first few lines:

```{r}
head(gapminder)
```

Another very helpful function for looking at your data is the `unique`()
function, to see the unique values in a particular column:

```{r}
unique(gapminder$continent)
```
