---
title: "Exploring Data Frames"
teaching: 20
exercises: 10
questions: "How can I manipulate a data frame?"
objectives:
- "Display basic properties of data frames including size and class of the columns, names, and first few rows."
- "Add and remove rows or columns."
- "Remove rows with `NA` values."
- "Append two data frames."
- "Understand what a `factor` is."
- "Convert a `factor` to a `character` vector and vice versa."
keypoints:
- "Use `str()`, `summary()`, `nrow()`, `ncol()`, `dim()`, `colnames()`, `rownames()`, `head()`, and `typeof()` to understand the structure of a data frame."
- "Understand the basics of subsetting data frames using `[]` and `$`"
- "Read in a csv file using `readr::read_csv()`."
source: Rmd
---

# Exploring Data Frames

```{r, include=FALSE}
source("bin/chunk-options.R")
knitr_fig_path("05-")
```

At this point, you've seen it all: in the last lesson, we toured all the basic
data types and data structures in R. Everything you do will be a manipulation of
those tools. But most of the time, the star of the show is the data frame—the table that we created by loading information from a csv file. In this lesson, we'll learn a few more things
about working with data frames.

```{r}
library(readr)
cats <- read_csv("data/feline-data.csv")
```

We can find the size and various other attributes of our data with a few handy functions:

```{r}
nrow(cats)
ncol(cats)
dim(cats)
```

```{r}
names(cats)
str(cats)
```

A way to get a nice overall view of a data frame is to use the `summary` function. This function can be used on various objects in R. For data frames, `summary` yields a numeric, tabular, or descriptive summary of each column. Factor columns are summarized by the number of items in each level, numeric or integer columns by the descriptive statistics (quartiles and mean), and character columns by its length, class, and mode.  

```{r}
summary(cats)
```

A really nice feature of RStudio is that it allows us to view our data in a 
very readable and familiar way. Go ahead and click on the name of the `cats`
data frame in the environment pane.

Note that this is a read-only view, it's not editable (which is a good thing!)

## Introducing the Gapminder data

So far, you have seen the basics of manipulating data frames with our cat data;
now let's use those skills to digest a more realistic dataset. Let's read in the
`gapminder` dataset that we downloaded previously:

### Challenge 1
>
> Read in the gapminder dataset using read_csv and assign it to a data frame
> called `gapminder`:
>
> ```{r}
> gapminder <- read_csv("data/gapminder_data.csv")
> ```
>
> How many rows and columns does `gapminder` have?
>
> What kind of data is in the `pop` and `continent` columns?
>
> <details>
> 
> <summary>
> Solution to challenge 1
> </summary>
> 
> <br />
> 
> ```{r}
> dim(gapminder)
> nrow(gapminder)
> ncol(gapminder)
> str(gapminder)
> summary(gapminder)
> ```
> </details>

The first thing we should always do is check out what the data looks like with
`str` or `summary`:

```{r}
str(gapminder)
```


```{r}
summary(gapminder$country)
```

::: {.rmdtip}

**Miscellaneous Tips**

* Files can also be downloaded directly from the Internet into a local
folder of your choice onto your computer using the `download.file` function.
The `read_csv` function can then be executed to read the downloaded file from the download location, for example,

```{r eval=FALSE, echo=TRUE}
download.file("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv", destfile = "data/gapminder_data.csv")
gapminder <- read_csv("data/gapminder_data.csv")
```

* Alternatively, you can also read in files directly into R from the Internet by replacing the file paths with a web address in `read_csv`. One should note that in doing this no local copy of the csv file is first saved onto your computer. For example,

```{r eval=FALSE, echo=TRUE}
gapminder <- read_csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv")
```

* You can read directly from excel spreadsheets without
converting them to plain text first by using the [readxl](https://cran.r-project.org/package=readxl) package.

:::

We'll also likely want to know what the titles of all the columns are, so we can
ask for them:

```{r}
names(gapminder)
```

At this stage, it's important to ask ourselves if the structure R is reporting
matches our intuition or expectations; do the basic data types reported for each
column make sense? If not, we need to sort any problems out now before they turn
into bad surprises down the road, using what we've learned about how R
interprets data, and the importance of *strict consistency* in how we record our
data.

Once we're happy that the data types and structures seem reasonable, it's time
to start digging into our data proper. Check out the first few lines:

```{r}
head(gapminder)
```



Another very helpful function for looking at your data is the `unique` 
function, to see the unique values in a particular column:

```{r}
unique(gapminder$continent)
```


To make sure our analysis is reproducible, we should put the code
into a script file so we can come back to it later.

### Challenge 2

> Go to file -> new file -> R script, and write an R script
> to load in the gapminder dataset.

## Basic subsetting

Data frames are two dimensions - rows and columns. We can access these using a 
specific notation with square brackets.

```{r, eval=FALSE}
my_data_frame[rows, columns]
```

### Selecting rows

We can select rows using the square brackets like so:

```{r}
gapminder[1, ]
gapminder[100:110, ]
gapminder[c(1, 3, 5, 7, 9), ]
```

### Challenge 3
>
> It's good practice to also check the last few lines of your data and some in the middle. How would you do this?
>
>
> <details>
> 
> <summary>
> Solution to challenge 3
> </summary>
> 
> Use your knowledge of the dataset to choose some rows in the middle and at the end. 
>```{r}
>nrow(gapminder)
>gapminder[c(700, 750, 800),]
>gapminder[1700:1704,]
>```
>
>
> <br />
>To check the last few lines it's relatively simple as R already has a function for this:
>
>```{r}
>tail(gapminder)
>tail(gapminder, n = 15)
>```
></details>

Searching for ones specifically in the middle isn't too hard but we could simply ask for a few lines at random. How would you code this?

::: {.rmdtip}

**Tip: There are several ways to achieve this.**
The solution here presents one form of using nested functions, i.e. a function passed as an argument to another function. This might sound like a new ept, but you are already using it!

Remember `my_dataframe[rows, cols]` will print to screen your data frame with the number of rows and columns you asked for (although you might have d for a range or named columns for example). How would you get the last row if you don't know how many rows your data frame has? R has a function for . What about getting a (pseudorandom) sample? R also has a function for this.

~~~
gapminder[sample(nrow(gapminder), 5), ]
~~~

:::

It's unlikely that you'll select rows directly though, you'll usually want 
to use some criteria. Say we just wanted the rows from Asia:

```{r}
romania <- gapminder[gapminder$continent == "Asia", ]
romania
```

This works because subsetting can take either numerical positions like we showed
above, OR a vector of `TRUE` and `FALSE`, returning the rows that are `TRUE`.
`gapminder$contintent == "Asia"` generates a vector of `TRUE` and `FALSE`
values:

```{r}
gapminder$continent == "Asia"
```

and inserting that vector into the first position of the square brackets
gives us just the rows where the `contintent` column is `"Asia"`.

Another very useful operator in this context is the `%in%` operator - it can 
tell you if a value is _one of_ a set of values:

Say we wanted to select just Canada and Mexico:

```{r}
gapminder$country %in% c("Canada", "Mexico")
```

Now we can insert this into the first position of the square brackets:

```{r}
can_mex <- gapminder$country %in% c("Canada", "Mexico")
gapminder[can_mex, ]
```

### Challenge 4
> Select all the rows in `gapminder` with data from 1975 and earlier
>
> <details>
> 
> <summary>
> Solution to challenge 4
> </summary>
> 
> <br />
> ```{r}
> gapminder[gapminder$year <= 1975, ]
> ```
> </details>

We saw previously how we can select a single column using the dollar sign `$`:

```{r}
gapminder$country
```

If we wanted to select a subset of columns, we can specify that in the second 
position using a character vector of column names:

```{r}
gapminder[ , c("country", "year", "lifeExp")]
```

### Challenge 5
> Select all the rows in `gapminder` with data from Canada, and just the columns
> year, pop, and gdpPercap
> <details>
> 
> <summary>
> Solution to challenge 5
> </summary>
> 
> <br />
> ```{r}
> gapminder[gapminder$country == "Canada", c("year", "pop", "gdpPercap")]
> ```
> </details>
>

