---
title: "Subsetting Data Frames"
teaching: 20
exercises: 10
questions: "How can I manipulate a data frame?"
objectives:
- "Add and remove rows or columns"
- "Or access a single row or column"
- "Remove rows with `NA` values"
- "Subset a data frame by character value"
keypoints:
- "Understand the basics of subsetting data frames using `[]` and `$`"
- "Read in a csv file using `readr::read_csv()`."
source: Rmd
---

```{r, include=FALSE}
source("bin/chunk-options.R")
knitr_fig_path("08-")
```

# (PART) Beyond the Fundamentals {-} 

# Subsetting Data Frames 

Earlier in the course, you saw the basics of manipulating data frames with our gapminder data;
now let's add to those skills and learn to subset data frames using row and column positions. Let's read in the
gapminder dataset that we downloaded previously and assign it to a data frame called gapminder:

```{r}
 library(readr)
 gapminder <- read_csv("data/gapminder_data.csv")
```

## Basic subsetting

Data frames are two dimensions - rows and columns. We can access these using a 
specific notation with square brackets.

```{r, eval=FALSE}
my_data_frame[rows, columns]
```

### Selecting rows

We can select rows using the square brackets like so:

```{r}
gapminder[1, ]
gapminder[100:110, ]
gapminder[c(1, 3, 5, 7, 9), ]
```

### Challenge 2 (5 minutes)
>
> It's good practice to also check the last few lines of your data and some in the middle. How would you do this?
>
>
> <details>
> 
> <summary>
> Solution to challenge 2
> </summary>
> 
> Use your knowledge of the dataset to choose some rows in the middle and at the end. 
>```{r}
>nrow(gapminder)
>gapminder[c(700, 750, 800),]
>gapminder[1700:1704,]
>```
>
>
> <br />
>To check the last few lines it's relatively simple as R already has a function for this:
>
>```{r}
>tail(gapminder)
>tail(gapminder, n = 15)
>```
></details>

Searching for ones specifically in the middle isn't too hard but we could simply ask for a few lines at random. How would you code this?

::: {.rmdtip}

**Tip: There are several ways to achieve this.**
The solution here presents one form of using nested functions, i.e. a function passed as an argument to another function. This might sound like a new ept, but you are already using it!

Remember `my_dataframe[rows, cols]` will print to screen your data frame with the number of rows and columns you asked for (although you might have d for a range or named columns for example). How would you get the last row if you don't know how many rows your data frame has? R has a function for . What about getting a (pseudorandom) sample? R also has a function for this.

~~~
gapminder[sample(nrow(gapminder), 5), ]
~~~

:::

It's unlikely that you'll select rows directly though, you'll usually want 
to use some criteria. Say we just wanted the rows from Asia:

```{r}
asia <- gapminder[gapminder$continent == "Asia", ]
asia
```

This works because subsetting can take either numerical positions like we showed
above, OR a vector of `TRUE` and `FALSE`, returning the rows that are `TRUE`.
`gapminder$contintent == "Asia"` generates a vector of `TRUE` and `FALSE`
values:

```{r}
gapminder$continent == "Asia"
```

and inserting that vector into the first position of the square brackets
gives us just the rows where the `contintent` column is `"Asia"`.

Another very useful operator in this context is the `%in%` operator - it can 
tell you if a value is _one of_ a set of values:

Say we wanted to select just Canada and Mexico:

```{r}
gapminder$country %in% c("Canada", "Mexico")
```

Now we can insert this into the first position of the square brackets:

```{r}
can_mex <- gapminder$country %in% c("Canada", "Mexico")
gapminder[can_mex, ]
```

We saw previously how we can select a single column using the dollar sign `$`:

```{r}
gapminder$country
```

If we wanted to select a subset of columns, we can specify that in the second 
position using a character vector of column names:

```{r}
gapminder[ , c("country", "year", "lifeExp")]
```


### Challenge 3 (5 minutes)
> Select all the rows in `gapminder` with data from 1975 and earlier
>
> <details>
> 
> <summary>
> Solution to challenge 3
> </summary>
> 
> <br />
> ```{r}
> gapminder[gapminder$year <= 1975, ]
> ```
> </details>


### Challenge 4 (5 minutes)

> Select all the rows in `gapminder` with data from Canada, and just the columns
> year, pop, and gdpPercap

> <details>
> 
> <summary>
> Solution to challenge 4
> </summary>
> 
> <br />
> ```{r}
> gapminder[gapminder$country == "Canada", c("year", "pop", "gdpPercap")]
> ```
> </details>
>

