---
title: "Intro to bcdata"
teaching: 30
exercises: 15
questions:
- "How can I pull data directly from the B.C. Data Catalogue using R?"
objectives:
- "Understand the value of importing data programatically for more auditable and reproducible work flows"
- "Become familiar with the bcdata R package"
- "Learn how to use bcdata with dplyr and ggplot2 packages"
source: Rmd
---

```{r, include=FALSE}
source("bin/chunk-options.R")
knitr_fig_path("09-")
```


# Introduction to the bcdata package

The bcdata R package contains functions for searching & retrieving data directly from the B.C. Data Catalogue.

```{r}
library(bcdata)
```

Let's look in the B.C. Data Catalogue for some data on libraries in British Columbia:

```{r}
bcdc_search("public libraries")
```

There are many catalogue records with data relating to libraries. Let's examine the first returned record more closely:


```{r}
bcdc_get_record("bc-public-libraries-systems-branches-and-locations")
```

Next, we can use the `bcdc_get_data` function to load this data directly into R:

```{r}
bcdc_get_data(record = "bc-public-libraries-systems-branches-and-locations")

lib_raw <- bcdc_get_data(record = 'bc-public-libraries-systems-branches-and-locations',
                          resource = 'ed17f111-fb39-46b3-89aa-5375592fbb01')
```


> ## Challenge 1 (5 minutes)
>
> Take a few minutes to search the B.C. Data Catalogue for data sets for a different topic. Identify and import one tabular data set into your R session > (e.g. a CSV or XLS resource).
>


Let's return to our libraries data. As we learned earlier in the workshop, a useful first step is to get familiar with the data.

```{r}
str(lib_raw)
names(lib_raw)
```

Let's tidy up this data frame and find out how many books&mdash;physical materials or volumes&mdash;are available within each City:

```{r}
library(dplyr)

lib_tidy <- lib_raw %>%
  select(`Library System`, Location, City, num_books = "460x Total branch physical materials, volumes held")


sum_books <- lib_tidy %>% 
  group_by(City, `Library System`) %>% 
  summarize(total_books = sum(num_books))
```

> ## Challenge 2 (10 minutes)
>
> Using `dplyr`, determine the total number of books within each Library System in British Columbia.
>
>
>```{r}
> sum_books %>% 
>  group_by(`Library System`) %>% 
>  summarize(totals = sum(total_books))
>```


Now that we have imported, tidied, and analyzed our library data, let's make a static data visualization to share and communicate the results with others.

```{r}
library(ggplot2)

sum_books %>% 
  ggplot(aes(total_books, City)) +
  geom_col()

sum_books %>% 
  filter(total_books > 100000) %>% 
  ggplot(aes(total_books, City)) +
  geom_col()
```

> ## Challenge 3 (10 minutes)
>
> Iterate the above plot to include information on the number of books by Library Service. Add a title to the plot. Try other design changes to make the plot more readable.
>
>
>
>```{r}
>library(ggplot2)
>
>sum_books %>% 
>  filter(total_books > 100000) %>% 
>  ggplot(aes(total_books, City)) +
>  geom_col() +
>  facet_wrap(vars(`Library System`)) +
>  theme_minimal() +
>  labs(x = NULL,
>       y = NULL,
>       title = "Total Number of Books by City & Library System in B.C.",
>       caption = "Data sourced from the B.C. Data Catalogue")
>
>```



```{r}
lib_spatial <- bcdc_query_geodata("bc-public-libraries-systems-branches-and-locations") %>%
  collect()

mapview::mapview(lib_spatial)

lib_spatial %>% 
  ggplot() + 
  geom_sf()

lib_spatial %>% 
  left_join(lib_tidy,
            by = c("SERVICE_POINT_NAME" = "Location")) %>% 
  ggplot() +
  geom_sf(aes(size = num_books), colour = "blue") +
  geom_sf(data = bcmaps::bc_bound(), fill = NA) +
  theme_minimal()

```


