---
title: "Intro to bcdata"
teaching: 30
exercises: 30
questions:
- "How can I import data directly from the B.C. Data Catalogue using R?"
objectives:
- "Understand the value of importing data programatically for more auditable and reproducible work flows"
- "Become familiar with the bcdata R package"
- "Learn how to use bcdata with the dplyr and ggplot2 packages"
source: Rmd
---

```{r, include=FALSE}
source("bin/chunk-options.R")
knitr_fig_path("09-")
```


# Introduction to the bcdata package


The British Columbia government hosts over 2000 tabular and geospatial data sets
in the B.C. Data Catalogue. Most data is available through the B.C. Data
Catalogue under an open licence. The `bcdata` package provides programming
interfaces the B.C. Data Catalogue.  This allows R users to search, download and
import metadata and data from the B.C. Data Catalogue into their R session.
Through this functionality the `bcdata` package connects British Columbia
government public data holdings in the B.C. Data Catalogue with the vast
capabilities of R.


## Getting started

First we need to install the `bcdata` package:

```{r, eval=FALSE}
install.packages("bcdata")
```

And next we need to load the `bcdata` package, along with `dplyr` and `ggplot2` which we installed and used in earlier lessons:

```{r, message=FALSE}
library(bcdata)
library(dplyr)
library(ggplot2)
```

## Searching the B.C. Data Catalogue

Let's use the `bcdata` package to search the B.C. Data Catalogue for some data on _**public libraries**_ in British Columbia:

```{r}
bcdc_search("public libraries", n = 5)
```

There are many catalogue records with data relating to libraries. Let's examine the first returned record more closely:

```{r}
bcdc_get_record("bc-public-libraries-systems-branches-and-locations")
```

## Importing data from the B.C. Data Catalogue

Next, we can use the `bcdc_get_data()` function to import this data directly into R: 

```{r, eval=FALSE}
bcdc_get_data(record = "bc-public-libraries-systems-branches-and-locations")
```


The _bc-public-libraries-systems-branches-and-locations_ catalogue record
contains multiple data resources or files. If we run the above code
interactively, R will prompt you in the console to select one of the available
resources, as it does not know which one you want to download. The full code to
import the resource from the record&mdash;using the record's permanent id to
guard against future name changes&mdash;is provided to the user for next time:

```{r}
lib_raw <- bcdc_get_data(record = '3d2318d4-8f5d-4208-88f5-995420d7c58f',
                         resource = 'ed17f111-fb39-46b3-89aa-5375592fbb01')
```

### Challenge 1 (5 minutes)
>
> Take a few minutes to search the B.C. Data Catalogue for data sets for a different topic. Identify and import one tabular data set into your R session (e.g. a CSV or XLS resource). Entering `bcdc_browse()` into the R console will open the catalogue page in your default web browser. 


## Practice manipulating data with `dplyr`  


Let's return to the libraries data frame `lib_raw` we imported. As we learned
earlier in the workshop, a useful first step is to get familiar with the data:

```{r}
str(lib_raw)
names(lib_raw)
```

Let's tidy up this data frame and find out how many books&mdash;physical
materials or volumes&mdash;are available between _51 and 53 degrees Latitude_. Remember
that we previously used `select()` to return only columns that we wanted. Here we
are _also_ using `select()` to rename `460x Total branch physical materials, volumes held`
to `num_books`:

```{r}
library(dplyr)

lib_tidy <- lib_raw %>%
  filter(Latitude <= 53, Latitude >= 51) %>% 
  select(`Library System`, Location, City, num_books = `460x Total branch physical materials, volumes held`)


sum_books <- lib_tidy %>% 
  group_by(City, `Library System`) %>% 
  summarize(total_books = sum(num_books))
```

### Challenge 2 (10 minutes)
>
> Using `dplyr`, determine the total number of books within each Library System between 51 and 53 degrees latitude in British Columbia.
>
>
> <details>
> 
> <summary>
> Solution to challenge 2
> </summary>
> 
> <br />
>```{r}
> sum_books %>% 
>  group_by(`Library System`) %>% 
>  summarize(totals = sum(total_books))
>```
> </details>



## Practice plotting data with `ggplot2`  

Now that we have imported, tidied, and summarized our library data, let's make a static data visualization to share and communicate the results with others.

```{r }
library(ggplot2)

sum_books %>% 
  ggplot(aes(x = total_books, y = City)) +
  geom_col()
```


### Challenge 3 (10 minutes)

> Iterate the above plot to include information on the number of books by Library Service. Add a title to the plot. Try other design changes to make the plot more readable. 
>
>
> <details>
> 
> <summary>
> Solution to challenge 3
> </summary>
> 
> <br />
>```{r}
>library(ggplot2)
>
>sum_books %>% 
>  ggplot(aes(x = total_books, y = City, fill = `Library System`)) +
>  geom_col() +
>  theme_minimal() +
>  labs(x = NULL,
>       y = NULL,
>       title = "Total Number of Books by City & Library System in B.C.",
>       caption = "Data sourced from the B.C. Data Catalogue")
>
>```
> </details>

## Importing data from the B.C. Data Catalogue - Another Example
Let's try another example with `bcdata`. This time we are going to try importing
some data that is a bit messier. We are going to work with [_New Apprenticeship
Registrations by Fiscal Year and
Gender_](https://catalogue.data.gov.bc.ca/dataset/87f26989-ccde-407e-a500-5f0c304c460d)
data. This example will require you to reference elements you've learned
throughout the workshop.

You can view the data record we are going to be working with in your browser by
running the following command:

```{r eval=FALSE}
bcdc_browse("new-apprenticeship-registrations-by-fiscal-year-and-gender")
```

As we've previously learned we can use `bcdc_get_data()` to load data from the B.C. Data Catalogue directly into R. Let's try that now:

```{r}
new_reg <- bcdc_get_data(record = "new-apprenticeship-registrations-by-fiscal-year-and-gender")
new_reg
```

Well this doesn't look right. We can see near the bottom of our output that
there are the words of `Gender` and `Trade` so we know the data is there but we also have this mess of `<NA>`.
In this scenario it can be helpful to look at the raw data. Navigate back to the
catalogue page you opened earlier and click the <kbd>View</kbd> button. This
will take you to the
[Resource](https://catalogue.data.gov.bc.ca/dataset/87f26989-ccde-407e-a500-5f0c304c460d/resource/8f4f1f19-9027-4d0b-a686-d820bccddd16)
page. Click the link under **Resource** and the data will open up in Excel. The
first thing we notice is that the data does not actually start until 10 rows
down. `bcdc_get_data()` use `read_csv()` to read csv files and that function
conveniently has an argument called `skip` which allows us to specify the number
of lines to skip before reading the data. We can use it with `bcdc_get_data()`
like this:

```{r}
new_reg <- bcdc_get_data(
  record = "new-apprenticeship-registrations-by-fiscal-year-and-gender", 
  skip = 9
  )
new_reg
```

Ok this is looking _much_ better. We now have data that is rectangular and looks
more like something we can work with. We do have one more thing to take
care of though. You will notice the data is filled with `...` values. Those are missing
values in the data. R has a specific way of handling missing values and we need
to tell `bcdc_get_data()` (via `read_csv()`) that those are missing values. We
do this by adding another argument so that R interprets the `...` as missing values:

```{r}
new_reg <- bcdc_get_data(
  "new-apprenticeship-registrations-by-fiscal-year-and-gender", 
  skip = 9, 
  na = "..."
  )
new_reg
```

Now we are ready to work with this data. One thing you might notice is that this
data is in _wide_ format. The number of individuals of a given `Gender` and
`Trade` are repeated over time across _columns_ with each column being a fiscal
year. This is great for making a nice table but less great if we want to make a
plot. To reshape the data into long format, we can again make use of
`pivot_longer` from the `tidyr` package which we covered in the [tidyr section][Create Tidy Data with tidyr]. We only want to pivot the fiscal years
so we need to specify those columns. In addition, we want to give our two new
columns so descriptive names:

```{r}
library(tidyr)

new_reg_long <- new_reg %>% 
  pivot_longer(
    cols = `FY2011/12`:`FY2020/21`,
    names_to = "fiscal_year",
    values_to = "registrations"
  )
new_reg_long
```

There is actually quite a bit of data here while likely represents the number
possible values for `Trade`. We can check the number of unique values in the
`Trade` column using a handy function from `dplyr` called `n_distinct()`:

```{r}
n_distinct(new_reg_long$Trade)
```

Sure enough there are `r n_distinct(new_reg_long$Trade)` of them! Similarly, let's have a look at what values are in the `Gender` column using the `unique()` function you learned earlier:

```{r}
unique(new_reg_long$Gender)
```

To simplify things let's take a subset of the Trades in the data and only look at Men and Women:
```{r}
new_reg_long <- new_reg %>% 
  pivot_longer(
    cols = `FY2011/12`:`FY2020/21`,
    names_to = "fiscal_year",
    values_to = "registrations", 
    values_drop_na = TRUE
  ) %>% 
  filter(
    Gender %in% c("Men", "Women"),
    Trade %in% c("Boilermaker", "Climbing Arborist", "Tilesetter", "Landscape Horticulturist", "Baker")
    )
```

With this tidied and filtered dataset, we now have something that we can plot
using the skills we've learned in the [ggplot2 section][Creating Publication-Quality Graphics with ggplot2]. One thing to note is that we made
use of the `ncol` argument in `facet_wrap()` so that the faceted plots would
share the x-axis:

```{r}
library(ggplot2)

ggplot(new_reg_long, aes(x = fiscal_year, y = registrations, fill = Trade)) +
  geom_col() +
  facet_wrap(vars(Gender), ncol = 1)
```

Try customizing the output with different `Trade` categories or different ggplot2 customizations. For example what strategies might you take to tidy up the x-axis?


### Challenge 4 (15 minutes)
>
> Navigate to the [BC Data Catalogue](https://catalogue.data.gov.bc.ca/datasets) and choose a dataset that has a `Resource Storage Format` that is `csv` or `xlsx`. An example of this is the [BC Seafood Production data](https://catalogue.data.gov.bc.ca/dataset/bc-seafood-production-2018-2020). You can use this dataset or one of your choosing. Import your data into R and make a plot, any plot at all. Try to apply the skills you've learned in this workshop to make a pretty plot. 
> 
> <details>
> 
> <summary>
> Solution to challenge 4
> </summary>
> 
> <br />
>```{r}
>seafood_prod <- bcdc_get_data("e591a756-6b99-4986-b8ce-f341bd47228f", resource = '9cc32dc4-bd9a-4baf-8210-770ac62d7985')
>
>seafood_by_year <- seafood_prod %>% 
>  filter(Year == 2020) %>% 
>  group_by(`Species/Product`) %>% 
>  summarise(value = sum(`Wholesale Value ($ million)`)) 
>
>seafood_by_year %>% 
>  ggplot(aes(y = `Species/Product`, x = value, fill = `Species/Product`)) +
>  geom_col() +
>  scale_fill_viridis_d() +
>  scale_x_continuous(labels = scales::dollar) +
>  guides(fill = "none") +
>  theme_minimal() +
>  labs(title = "B.C. Seafood Production - 2020", 
>       x = "Wholesale Value ($ million)",
>       caption = "https://catalogue.data.gov.bc.ca/dataset/e591a756-6b99-4986-b8ce-f341bd47228f") +
>  theme(axis.title.y = element_blank())
>```
> </details>


