---
title: "Intro to bcdata"
teaching: 30
exercises: 30
questions:
- "How can I import data directly from the B.C. Data Catalogue using R?"
objectives:
- "Understand the value of importing data programatically for more auditable and reproducible work flows"
- "Become familiar with the bcdata R package"
- "Learn how to use bcdata with the dplyr and ggplot2 packages"
source: Rmd
---

```{r, include=FALSE}
source("bin/chunk-options.R")
knitr_fig_path("09-")
```


# Introduction to the bcdata package


The British Columbia government hosts over 2000 tabular and geospatial data sets
in the B.C. Data Catalogue. Most data is available through the B.C. Data
Catalogue under an open licence. The `bcdata` package provides programming
interfaces the B.C. Data Catalogue.  This allows R users to search, download and
import metadata and data from the B.C. Data Catalogue into their R session.
Through this functionality the `bcdata` package connects British Columbia
government public data holdings in the B.C. Data Catalogue with the vast
capabilities of R.


## Getting started

First we need to install the `bcdata` package:

```{r, eval=FALSE}
install.packages("bcdata")
```

And next we need to load the `bcdata` package, along with `dplyr` and `ggplot2` which we installed and used in earlier lessons:

```{r, message=FALSE}
library(bcdata)
library(dplyr)
library(ggplot2)
```

## Searching the B.C. Data Catalogue

Let's use the `bcdata` package to search the B.C. Data Catalogue for some data on _**public libraries**_ in British Columbia:

```{r}
bcdc_search("public libraries")
```

There are many catalogue records with data relating to libraries. Let's examine the first returned record more closely:

```{r}
bcdc_get_record("bc-public-libraries-systems-branches-and-locations")
```

## Importing data from the B.C. Data Catalogue

Next, we can use the `bcdc_get_data` function to import this data directly into R: 

```{r, eval=FALSE}
bcdc_get_data(record = "bc-public-libraries-systems-branches-and-locations")
```


The _bc-public-libraries-systems-branches-and-locations_ catalogue record contains multiple data resources or files. If we run the above code interactively, R will prompt you in the console to select one of the available resources, as it does not know which one you want to download. The full code to import the resource from the record&mdash;using the record's permanent id to guard against future name changesname&mdash;is provided to the user for next time:

```{r}
lib_raw <- bcdc_get_data('3d2318d4-8f5d-4208-88f5-995420d7c58f',
                         resource = 'ed17f111-fb39-46b3-89aa-5375592fbb01')
```

> ## Challenge 1 (5 minutes)
>
> Take a few minutes to search the B.C. Data Catalogue for data sets for a different topic. Identify and import one tabular data set into your R session > (e.g. a CSV or XLS resource). Entering `bcdc_browse()` into the R console will open the catalogue page in your default web browser. 
>

## Practice manipulating data with `dplyr`  


Let's return to the libraries data frame `lib_raw` we imported. As we learned earlier in the workshop, a useful first step is to get familiar with the data:

```{r}
str(lib_raw)
names(lib_raw)
```

Let's tidy up this data frame and find out how many books&mdash;physical materials or volumes&mdash;are available within each City:

```{r}
library(dplyr)

lib_tidy <- lib_raw %>%
  select(`Library System`, Location, City, num_books = "460x Total branch physical materials, volumes held")


sum_books <- lib_tidy %>% 
  group_by(City, `Library System`) %>% 
  summarize(total_books = sum(num_books))
```

> ## Challenge 2 (10 minutes)
>
> Using `dplyr`, determine the total number of books within each Library System in British Columbia.
>
>
> >## Solution Challenge 2 (10 minutes)
> >```{r}
> > sum_books %>% 
> >  group_by(`Library System`) %>% 
> >  summarize(totals = sum(total_books))
> >```


## Practice plotting data with `ggplot2`  

Now that we have imported, tidied, and summarized our library data, let's make a static data visualization to share and communicate the results with others.

```{r}
library(ggplot2)

sum_books %>% 
  ggplot(aes(total_books, City)) +
  geom_col()

sum_books %>% 
  filter(total_books > 100000) %>% 
  ggplot(aes(total_books, City)) +
  geom_col()
```

> ## Challenge 3 (10 minutes)
>
> Iterate the above plot to include information on the number of books by Library Service. Add a title to the plot. Try other design changes to make the plot more readable.
>
>
> >## Solution Challenge 3 (10 minutes)
> >```{r}
> >library(ggplot2)
> >
> >sum_books %>% 
> >  filter(total_books > 100000) %>% 
> >  ggplot(aes(total_books, City)) +
> >  geom_col() +
> >  facet_wrap(vars(`Library System`)) +
> >  theme_minimal() +
> >  labs(x = NULL,
> >       y = NULL,
> >       title = "Total Number of Books by City & Library System in B.C.",
> >       caption = "Data sourced from the B.C. Data Catalogue")
> >
> >```

> ## Challenge 4 (20 minutes)
>
> Navigate to the [BC Data Catalogue](https://catalogue.data.gov.bc.ca/datasets) and choose a dataset that has a `Resource Storage Format` that is `csv` or `xlsx`. > An example of this is the [BC Seafood Production data](https://catalogue.data.gov.bc.ca/dataset/bc-seafood-production-2018-2020). You can use this dataset or one > of your choosing. Import your data into R and make a plot, any plot at all. Try to apply the skills you've learned in this workshop to make a pretty plot. 
> 
> >## Solution to Challenge 4 
> >```{r}
> >seafood_prod <- bcdc_get_data("e591a756-6b99-4986-b8ce-f341bd47228f", resource = '9cc32dc4-bd9a-4baf-8210-770ac62d7985')
> >
> >seafood_by_year <- seafood_prod %>% 
> >  filter(Year == 2020) %>% 
> >  group_by(`Species/Product`) %>% 
> >  summarise(value = sum(`Wholesale Value ($ million)`)) 
> >
> >seafood_by_year %>% 
> >  ggplot(aes(y = `Species/Product`, x = value, fill = `Species/Product`)) +
> >  geom_col() +
> >  scale_fill_viridis_d() +
> >  scale_x_continuous(labels = scales::dollar) +
> >  guides(fill = "none") +
> >  theme_minimal() +
> >  labs(title = "B.C. Seafood Production - 2020", 
> >       xlab = "Wholesale Value ($ million)",
> >       caption = "BC Seafood Production, 2018 - 2020: https://catalogue.data.gov.bc.ca/dataset/e591a756-6b99-4986-b8ce-f341bd47228f") +
> >  theme(axis.title.x = element_blank())
> >```

