---
title: Producing Reports With knitr Part 2
teaching: 30
exercises: 15
questions:
- "How can I integrate software and reports?"
objectives:
- Understand the value of writing reproducible reports
- Learn how to recognise and compile the basic components of an R Markdown file
- Become familiar with R code chunks, and understand their purpose, structure and options 
- Demonstrate the use of inline chunks for weaving R outputs into text blocks, for example when discussing the results of some calculations
- Be aware of alternative output formats to which an R Markdown file can be exported
keypoints:
- "Mix reporting written in R Markdown with software written in R."
- "Specify chunk options to control formatting."
- "Use `knitr` to convert these documents into PDF and other formats."
source: Rmd
---

```{r chunk_options, include=FALSE}
source("bin/chunk-options.R")
knitr_fig_path("15-")
```


 How things get compiled

When you press the "Knit" button, the R Markdown document is
processed by [`knitr`](http://yihui.name/knitr) and a plain Markdown
document is produced (as well as, potentially, a set of figure files): the R code is executed
and replaced by both the input and the output; if figures are
produced, links to those figures are included.

The Markdown and figure documents are then processed by the tool
[`pandoc`](http://pandoc.org/), which converts the Markdown file into an
html file, with the figures embedded.

```{r rmd_to_html_fig, fig.width=8, fig.height=3, fig.align="left", echo=FALSE}
par(mar=rep(0, 4), bty="n", cex=1.5)
plot(0, 0, type="n", xlab="", ylab="", xaxt="n", yaxt="n",
     xlim=c(0, 100), ylim=c(0, 100))
xw <- 10
yh <- 35
xm <- 12
ym <- 50
rect(xm-xw/2, ym-yh/2, xm+xw/2, ym+yh/2, lwd=2)
text(xm, ym, ".Rmd")

xm <- 50
ym <- 80
rect(xm-xw/2, ym-yh/2, xm+xw/2, ym+yh/2, lwd=2)
text(xm, ym, ".md")
xm <- 50; ym <- 25
for(i in c(2, 0, -2))
    rect(xm-xw/2+i, ym-yh/2+i, xm+xw/2+i, ym+yh/2+i, lwd=2,
         border="black", col="white")
text(xm-2, ym-2, "figs/")

xm <- 100-12
ym <- 50
rect(xm-xw/2, ym-yh/2, xm+xw/2, ym+yh/2, lwd=2)
text(xm, ym, ".html")

arrows(22, 50, 38, 50, lwd=2, col="slateblue", len=0.1)
text((22+38)/2, 60, "knitr", col="darkslateblue", cex=1.3)

arrows(62, 50, 78, 50, lwd=2, col="slateblue", len=0.1)
text((62+78)/2, 60, "pandoc", col="darkslateblue", cex=1.3)
```


 Chunk options

There are a variety of options to affect how the code chunks are
treated. Here are some examples:

- Use `echo=FALSE` to avoid having the code itself shown.
- Use `results="hide"` to avoid having any results printed.
- Use `eval=FALSE` to have the code shown but not evaluated.
- Use `warning=FALSE` and `message=FALSE` to hide any warnings or
  messages produced.
- Use `fig.height` and `fig.width` to control the size of the figures
  produced (in inches).

So you might write:

<pre>
&#96;&#96;&#96;{r load_libraries, echo=FALSE, message=FALSE}
library("dplyr")
library("ggplot2")
&#96;&#96;&#96;
</pre>

Often there will be particular options that you'll want to use
repeatedly; for this, you can set _global_ chunk options, like so:

<pre>
&#96;&#96;&#96;{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.path="Figs/", message=FALSE, warning=FALSE,
                      echo=FALSE, results="hide", fig.width=11)
&#96;&#96;&#96;
</pre>

The `fig.path` option defines where the figures will be saved. The `/`
here is really important; without it, the figures would be saved in
the standard place but just with names that begin with `Figs`.

If you have multiple R Markdown files in a common directory, you might
want to use `fig.path` to define separate prefixes for the figure file
names, like `fig.path="Figs/cleaning-"` and `fig.path="Figs/analysis-"`.


 Formatting figures 

We can define the location and size of plots using `fig.height`,`fig.width`
and `fig.align`. We can add captions using `fig.cap`, and many other parameters.



> ## Challenge 3
>
> Use chunk options to control the size of a figure and to hide the
> code. We can plot data from the ***faithful*** dataset, which is 
downloaded when you install R. Hint: you may want to first take a look 
at the data, then use the basic plot() command. 
> 
> > ## Solution to Challenge 3
> > 
> > <pre>
> > &#96;&#96;&#96;{r echo = FALSE, fig.width = 3, fig.align = 'centre'}
> > plot(faithful)
> > &#96;&#96;&#96;
> > </pre>
> > 
> > ```{r plot faithful, echo = FALSE, fig.width = 3, fig.align = 'centre'}
> > plot(faithful)
> > ```
> > 
> {: .solution}
{: .challenge}


 Finding help 

You can review all of the `R` chunk options by navigating to
the "R Markdown Cheat Sheet" under the "Cheatsheets" section 
of the "Help" field in the toolbar at the top of RStudio.


 Inline R code

You can make _every_ number in your report reproducible. Use
<code>&#96;r</code> and <code>&#96;</code> for an in-line code chunk,
like so: <code>&#96;r round(some_value, 2)&#96;</code>. The code will be
executed and replaced with the _value_ of the result.


For example; 
 <code>&#96;r round(3.141593, 2)&#96;</code> will show as 
 ``` `r round(3.141593, 2)` ```
 
We can also do other calculations on the fly within our Rmarkdown document. For example we include the number of instructors in this course and populate a sentence. Firstly lets use `length() ` to determine the number of instructors: 

<code>&#96; r length(c("Andy", "Gen", "Sam", "Steph"))&#96; </code>

when we run this in the R console we should get `4`.


Now we can add this as in-line code by using <code>&#96;r</code> and <code>&#96;</code> : 

<code> this course has &#96; r length(c("Andy", "Gen", "Sam", "Steph"))&#96; instructors </code>

The rendered fomat will look like 

``` this course has `r length(c("Sam", "Steph", "Andy", "Gen"))` instructors ```


Hint: Don't let these in-line chunks get split across lines.


Alternatively we can add a code chunk above our sentence which calculates and defines the variables. We can use `include=FALSE` , which will hide the code chunk but still evaluate the varibles. Note: `include=FALSE` is the same as `echo=FALSE` and `results="hide"`. 

<pre>
&#96;&#96;&#96;{r calculate instructors, include = FALSE }
no_instructors <- length(c("Andy", "Gen", "Sam", "Steph"))
&#96;&#96;&#96;
</pre>

```{r, include = FALSE }

no_instructors <- length(c("Andy", "Gen", "Sam", "Steph"))

```

<code> this course has &#96; r no_instructors &#96; instructors </code>

which will be rendered as: 

``` this course has `r no_instructors` instructors ```



> ## Challenge 4
>
> Try out a bit of in-line R code using a simple addition eg: 2+2. 
> 
> > ## Solution to Challenge 4
> > 
> > Here's some inline code to determine that 2 + 2 = `` `r
> > 2+2` ``.
> > 
> {: .solution}
{: .challenge}


 Other output options

You can also convert R Markdown to a PDF or a Word document. Click the
little triangle next to the "Knit" button to get a drop-down
menu. Or you could put `pdf_document` or `word_document` in the initial header
of the file.




> ## Tip: Creating PDF documents
>
> Creating .pdf documents may require installation of some extra software. If
> required this is detailed in an error message.
>
> - [TeX installers for Windows](https://miktex.org/2.9/setup).
> - [TeX installers for macOS](https://tug.org/mactex).
{: .callout}



 Resources

* [Knitr in a knutshell tutorial](http://kbroman.org/knitr_knutshell)
* [Dynamic Documents with R and knitr](http://www.amazon.com/exec/obidos/ASIN/1482203537/7210-20) (book)
* [R Markdown documentation](http://rmarkdown.rstudio.com)
* [R Markdown cheat sheet](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)
* [Getting started with R Markdown](https://www.rstudio.com/resources/webinars/getting-started-with-r-markdown/)
* [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/) (book by Rstudio team)
* [Reproducible Reporting](https://www.rstudio.com/resources/webinars/reproducible-reporting/)
* [The Ecosystem of R Markdown](https://www.rstudio.com/resources/webinars/the-ecosystem-of-r-markdown/)
* [Introducing Bookdown](https://www.rstudio.com/resources/webinars/introducing-bookdown/)
