---
title: "Introduction to R Workshop Challenges"
author: "DSP Data Science Team"
date: "14/02/2022"
output: 
  ioslides_presentation:
    css: ../bin/style-challenges.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("../bin/chunk-options-challenges.R")

library(readr)
library(ggplot2)

#data for ggplot challenges
gapminder <- read_csv("../data/gapminder_data.csv")
```

## Challenge 1 {#challenge-1}

Modify the example so that the figure shows how life expectancy has changed over time:

```{r, echo = TRUE}
 ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
 geom_point()
```

Hint: the gapminder dataset has a column called "year", which should appear on the x-axis.

## Solution to challenge 1 {#solution-1}

Here is one possible solution:

```{r ch1-sol}
ggplot(data = gapminder, mapping = aes(x = year, y = lifeExp)) +
geom_point()
```

## Challenge 2 {#challenge-2}

In the previous examples and challenge we've used the `aes` function to tell
the scatterplot **geom** about the **x** and **y** locations of each point.
Another *aesthetic* property we can modify is the point *colour*. 

Modify the code from the previous challenge to **colour** the points by the "continent"
column. 
 - What trends do you see in the data? 
 - Are they what you expected?

## Solution to challenge 2 {#solution-2}

```{r ch2-sol}
ggplot(data = gapminder, mapping = aes(x = year, y = lifeExp, colour = continent)) +
 geom_point()
```

## Challenge 3

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

