---
title: "Gapminder analyis"
author: "Andy Teucher"
date: "May 7, 2020"
output:
  word_document: default
---

```{r load, include=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(broom)

gapminder <- read_csv("data/gapminder_data.csv")

cntry <- "Netherlands"
```

## Analyzing GDP and Life Expectancy in `r cntry`


```{r filter-can, include = FALSE}
country_data <- filter(gapminder, country == cntry)
```

As we can see there is a relationship between GDP and Life Expectancy in `r cntry`.

```{r plot-1, echo=FALSE, message=FALSE}
ggplot(country_data, aes(y = lifeExp, x = gdpPercap)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

The output of our model, which is specified as **lifeExp = b0 + b1*gdpPercap + e** is:


```{r model, echo=FALSE}
country_data_lm <- lm(lifeExp ~ gdpPercap, data = country_data)
summary(country_data_lm)
```

The model diagnostics show that this is an adequate model that meets the 
assumptions of the regression:

```{r diagnostics, echo=FALSE}
plot(country_data_lm)
```



```{r broom, include=FALSE}
model_summary <- glance(country_data_lm)
model_summary
model_terms <- tidy(country_data_lm)
model_terms
model_predict <- augment(country_data_lm)
model_predict
```

```{r model-stats, echo=FALSE}
life_exp <- filter(model_terms, term == "gdpPercap")
life_exp_estimate <- round(life_exp$estimate, 5)
```

The predicted effect of per-capita GDP on life expectancy is `r life_exp_estimate`, 
meaning that for every dollar increase in GDP per-capita, our life expectancy 
increases by `r life_exp_estimate` years (See Figure 1).

```{r plot-ci, echo=FALSE, fig.cap=paste("Figure 1: Regression of GDP on Life Expectancy for", cntry)}
r2 <- round(model_summary$adj.r.squared, 2)

model_predict <- model_predict %>% 
  mutate(lower_cl = .fitted - 1.96*.se.fit, 
         upper_cl = .fitted + 1.96*.se.fit)

ggplot() +
  geom_ribbon(data = model_predict, 
              aes(x = gdpPercap, ymin = lower_cl, ymax = upper_cl), 
              alpha = 0.5, fill = "orange") + 
  geom_point(data = country_data, aes(x = gdpPercap, y = lifeExp)) + 
  geom_line(data = model_predict, aes(x = gdpPercap, y = .fitted)) + 
  geom_text(aes(x = 15000, y = 80), label = paste("R^2 =", r2))
```


